Parsing Postfix log files

### 2008-01-20 doc/logparser-acronyms.tex
Dealt with things which arose during my last meeting with Carl: * Improve the description of states in the emergent behaviour section. * Explained subclassing some more. * Added glossary and acronym entries for lmtp and virtual.

>>>>

\newacronym{LMTP}{Local Mail Transfer Protocol}{description={
    LMTP is a protocol derived from SMTP that removes the need for the
    server to maintain a mail delivery queue, instead relying on the client
    to maintain it.  Typically the client would be an MTA, and the server
    would be a delivery agent or a mail store.  Full details are available
    in \cite{lmtp-rfc}.
}}

### 2008-01-20 doc/logparser.tex
Dealt with things which arose during my last meeting with Carl: * Improve the description of states in the emergent behaviour section. * Explained subclassing some more. * Added glossary and acronym entries for lmtp and virtual.

Yet despite the brevity of the algorithm above the (simplified) flow chart
below contains twenty one states, three entry points, five exit points,
five loops, four explicit branches, and eight implicit branches.  Several
more states and corresponding branches are omitted for the sake of clarity
(these extra states are caused by solving the difficulties described in
section~\refwithpage{additional complications}).

The solution to the second last complication encountered
(section~\refwithpage{out of order log lines}) requires that a list of the
acceptable combinations of Postfix programs a mail can pass through;
however this does not correspond to the \textit{actions} a mail must pass
through.  Before that complication was overcome, i.e.\ for the majority of
the development of this parser, there was nothing in the parser explicitly
encoding the states or paths a mail could take; the vast majority of the
parser has no need for this flow of data to be specified.  This eases the
process of adding new actions (described in section~\refwithpage{adding new
actions}), as there is no requirement to explicitly insert the action into
an algorithmic version of the flow chart below.

@@

Adding new actions is almost as simple as adding new rules, though actions
are not stored in the database.  The implementor writes a subroutine which
accepts the standard arguments given to actions, and registers it as an
action by calling the parser method add\_actions() with the subroutine
name as a parameter.  No other work is required on the implementor's part;
all of their attention and effort can be focused on the correctness of
their action.  The only negative aspect is that the process involves
editing the parser source code, which makes upgrading to a later version of
the parser more difficult, though by no means impossible.  If the author of
the new action wishes they can take advantage of the parser's object
oriented implementation~\cite{Wikipedia-object-orientation} by subclassing
it and implementing their changes in the derived class, allowing future
upgrading of the parser with greatly reduced chance of conflicts.  The
action may need to extend the list of valid combinations described in
section~\refwithpage{out of order log lines} if the addition creates a
different set of acceptable programs, but this is unlikely to occur, as it
would require parsing log lines from Postfix components the parser
currently ignores.\footnote{The mail server used for development does not
utilise the \daemon{lmtp} or \daemon{virtual} delivery agents, so this
parser does not deal with those components.  Adding new rules to parse
their log lines is a simple process, though if their behaviour differs
significantly from the \daemon{smtp} or \daemon{local} delivery agents new
actions may be required.}

@@

The complications described in this section are listed in the order in
which they were encountered during development of the parser.  Each of
these complications caused the parser to operate incorrectly, generating
either warning messages or leaving mails in the state table.  The frequency
of occurrence is much higher at the start of the list, with the first
complication occurring several orders of magnitude more frequently than the
last.  When deciding which problem to address next, the most common was
always chosen, as resolving the most common problem would yield the biggest
improvement in the parser, and also prune the greatest number of mails from
the state tables and error messages.

@@

@@

>>>>

Yet despite the brevity of the algorithm above the (simplified) flow chart
below contains:

    \item Twenty one states (not including explicit branches).

    \item Three entry points (1, 2, 3).

    \item Five exit points (8, 14, 17, 20, 25).

    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).

    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
        by an action, determining what will happen next to a mail.

    \item Four implicit branches, where the transition is determined by the
        next log line which is processed for that mail, e.g.\ state 1 can
        transition to either state 4 or state 5.  (1--\{4,5\},
        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})

    Several
more states and corresponding branches are omitted for the sake of clarity
(these extra states are caused by solving the difficulties described in
section~\refwithpage{additional complications}).

The solution to the second last complication encountered
(section~\refwithpage{out of order log lines}) requires that a list of the
acceptable combinations of Postfix programs a mail can pass through;
however this does not correspond to the \textit{actions\/} a mail must pass
through.  Before that complication was overcome, i.e.\ for the majority of
the development of this parser, there was nothing in the parser explicitly
encoding the states or paths a mail could take; the vast majority of the
parser has no need for this flow of data to be specified.  This eases the
process of adding new actions (described in section~\refwithpage{adding new
actions}), as there is no requirement to explicitly insert the action into
an algorithmic version of the flow chart below.

@@

Adding new actions is almost as simple as adding new rules, though actions
are not stored in the database.  The implementor writes a subroutine which
accepts the standard arguments given to actions, and registers it as an
action by calling the parser subroutine add\_actions() with the name of the
new action subroutine as a parameter.  No other work is required on the
implementor's part; all of their attention and effort can be focused on the
correctness of their action.  The only negative aspect is that the process
involves editing the parser source code, which makes upgrading to a later
version of the parser more difficult, though by no means impossible.  If
the author of the new action wishes they can take advantage of the parser's
object oriented implementation~\cite{Wikipedia-object-orientation} by
subclassing it and implementing their changes in the derived class,
allowing future upgrading of the parser with greatly reduced chance of
conflicts.\footnote{The real difference between the two approaches is where
the new code is placed.  The simpler option is to change the parser code
directly, but those changes will then have to be made to subsequent
versions of the parser, and as the scope of the changes increases so does
the chance of conflict.  The more time consuming option is to write a
subclass containing the new actions and change the program which invokes
the parser so that it uses the subclass rather than the parser; the changes
required to the program invoking the parser are minor and much less likely
to lead to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser will
avail of the new functionality.} The action may need to extend the list of
valid combinations described in section~\refwithpage{out of order log
lines} if the addition creates a different set of acceptable programs, but
this is unlikely to occur, as it would require parsing log lines from
Postfix components the parser currently ignores.\footnote{The mail server
used for development does not utilise the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse their log lines is a
simple process, though if their behaviour differs significantly from the
\daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit is that the logs used exhibit
the idiosyncrasies and foibles a mail server in the wild must deal with,
but the downside is that significantly altering the configuration just to
log messages from a different Postfix component is not an option.}

@@

The complications described in this section are listed in the order in
which they were encountered during development of the parser.  Each of
these complications caused the parser to operate incorrectly, generating
either warning messages or leaving mails in the state table.  The frequency
of occurrence is much higher at the start of the list, with the first
complication occurring several orders of magnitude more frequently than the
last.  When deciding which problem to address next the most common was
always chosen, as resolving the most common problem would yield the biggest
improvement in the parser, and also prune the greatest number of mails from
the state tables and error messages.

@@

\glossary{name={lmtp},description={
    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
}}

@@

\glossary{name={virtual},description={
    Virtual is the Postfix component responsible for delivery of mails to
    virtual domains.  With \daemon{local} delivery the destination is
    determined only by the portion of the email address on the left side of
    the \at{}, whereas with \daemon{virtual} delivery the destination is
    determined by the entire email address, e.g.\ if the server considers
    itself responsible for both \textbf{example.org} and
    \textbf{example.net} domains: \daemon{local} considers
    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
    the same mailbox, whereas \daemon{virtual} considers them to have
    different mailboxes.  Virtual delivery is used when a server hosts
    multiple domains where a username may be present in more than one
    domain but represent different users in each.  The definitive
    documentation is
    \url{http://www.postfix.org/virtual.8.html}.
}}
